---
title: "General Framework for Seat Occupancy Expectation"
author:
  name: "Fukumoto Tomoya"
  affiliation: "Operation Planning"
date: "2018-08-20"
output: html_document
#abstract: "Maximizing seat occupancy is an important issue for theme park operations. We established a mathematical framework to predict seat occupancy of ride attractions. Our model is based on Markov chain theory. Since our model is general, we can apply it to any attraction."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction
One of the most important issues in theme park operation is increasing guest counts at attractions.
Since guest counts can be decomposed into dispatch count and seat occupancy, we focus on these two factors separately.
The aim of this study is to establish a mathematical framework to predict seat occupancy of ride attractions.
We revealed that seat occupancy can be derived from statistics of group sizes of arriving guests, the configuration of seats, queue formation, and grouping strategy.
This framework will help us to optimize queue formation and grouping strategy.
Because the model is general, that is, not dependent on any specific attraction, the framework can be applied to any attraction.

## Modeling and Framework
### Problem
For ride attractions, grouping operation is deeply related to seat occupancy.
Attraction crew members form a group to dispatch a vehicle from arriving guests without breaking up friend groups.
Such crew members are called groupers.
Guests arrive continuously.
Groupers guide guests who arrived earlier first.
Groupers cannot choose or predict the size of the next friend group to arrive.

Almost all vehicle seat arrangements are arranged in a rectangular pattern.
This means seats are arranged in rows, with multiple rows lined up.
The grouping operation involves seating guest groups in the same row.
Friend groups must remain together in the same row and sit consecutively.

The grouping problem is how to maximize guest experience while filling seats with guest groups.
Groupers ask the size of each group and select a row to guide them to.
The grouping strategy is to select an optimal row when a group of a given size arrives.

Therefore, there exists an upper bound on seat occupancy for each attraction.

The upper bound depends on statistics of group sizes of arriving guests and the pattern of seat configuration.
If we understand the upper bound for an attraction, we can evaluate the efficiency of its grouping operation and determine whether the seat occupancy can be improved.

### Mathematical Modeling
Our problem can be handled as a state-action-reward model.

- **state**:  the seats pattern filled by guests
- **action**: decision which row to seat arriving guests and whether to dispatch a vehicle
- **reward**: the number of guests who take a seat when a vehicle is dispatched

To handle this model, we define several variables for state and transition:

- $n$: the number of seats in a row (for multi-row configurations, $n_i$ denotes seats in row $i$)
- $x_i \in \mathbb{N}$: the size of the $i$-th arriving group
- $\Sigma$: the set of all possible seat occupancy states
- $\sigma \in \Sigma$: a seat occupancy state (pattern representing which seats are filled)
- $s_i \in \Sigma$: the seat occupancy state after the $i$-th group has been seated

After a vehicle is dispatched, a new empty vehicle arrives.
This means the state will be initialized as $\sigma_0$.

<!--a1[label = '<I>s@_{i}</I>', shape = circle];-->

```{r markov, fig.width = 5, fig.height = 2}
library(DiagrammeR)
grViz(
  "digraph dot{
    graph[
      rankdir = LR,
      label = 'Fig.1: State, Action, and Transition',
      fontsize = 15
    ]
    node[
      fixedsize = false
    ]
      a1[label = '<I>s</I>@_{<I>i</I>}=<I>&sigma;</I>', shape = circle];
      a2[label = 'Which row\n to seat?', shape = square];
    edge[label = '<I>x@_{i}</I>'] a1 -> a2 ;
      a3[label = 'Dispatch\n or not?', shape = square];
    edge[label = ''] a2 -> a3;
      a4[label = '<I>s</I>@_{<I>i</I>+1}=<I>&sigma;</I>&prime;', shape = circle];
      edge[label = ''] a3 -> a4;

}
  "
)
```

We assume that the situation is stationary.
This stationarity assumption means that the probability of a group of size $k$ arriving is independent of time, i.e., $P[x_i=k]=p_k$ for all $i$.

## Simple case: Markov chain
If the seat layout is a single row, we have only two options: seat the group or dispatch the vehicle, since we cannot choose which specific seats to provide.

In this single-row case with first-come-first-served seating, the state can be simplified to the number of filled seats.
If the row has $n$ seats, the state space becomes $\Sigma = \{0, 1, 2, \ldots, n\}$, where each state $\sigma \in \Sigma$ represents the number of filled seats.
Thus $s_i \in \{0, 1, \ldots, n\}$ denotes the number of filled seats after the $i$-th group has been seated.

Since friend groups must sit together in the same row and cannot be split, groups of size $k > n$ cannot be seated in a single-row configuration.
Therefore, $p_k = 0$ for all $k > n$ is a necessary consequence of this physical constraint.

In this single-row case, our model can be treated as a simple **Markov chain model**.

### Stationary state probability distribution
Under mild conditions, a finite-state Markov chain converges to a unique stationary distribution.
For our model with finite state space $\Sigma = \{0, 1, \ldots, n\}$ and typical arrival distributions, these conditions are satisfied, guaranteeing the existence of a stationary state.

We define the **stationary state probability vector** as:
$$
\pi = (\pi_0, \pi_1, \ldots, \pi_{n})^T
$$
where $\pi_j = \lim_{i \to \infty} P[s_i = j]$ for $j \in \{0, 1, \ldots, n\}$ represents the long-run probability that exactly $j$ seats are filled when a grouping decision is made.

Given the arrival distribution $p_k = P[x_i=k]$, we can derive the stationary distribution $\pi$ through the three steps outlined below.
Once $\pi$ is obtained, we can calculate the expected seat occupancy, which represents the theoretical upper bound under the FIFO constraint.

### Steps to solve the problem
We solve this in three steps:

1. Construct a transition matrix based on the number of seats and the distribution of group sizes.
2. Identify the stationary state.
3. Calculate the expected seat occupancy from the stationary probability distribution.

#### Setting a transition matrix
Suppose a group of size $k$ arrives.
If the number of empty seats is greater than or equal to $k$, the group is seated.
If it is less than $k$, the crew dispatches the current vehicle and the group is seated in a new vehicle.

- If $s_i + x_i \leq n$ then $s_{i+1}=s_i+x_i$ 
- If $s_i + x_i > n$ then $s_{i+1}=x_i$

This means the state probability can be described by the previous state and arrival distribution.
Since the arrival distribution is independent of the current state, the transition probabilities are
$$
\begin{eqnarray}
T_{jk}&=&P[s_{i+1}=k|s_{i}=j]\\
  &=& \sum_{x=1}^{n} p_x \left[ \mathbb{1}_ {[j+x=k, j+x \leq n]}
  + \mathbb{1}_ {[x=k, j+x > n]} \right]
\end{eqnarray}
$$
We call such a matrix the **transition matrix**.

We show an example with the simplest case: two seats.
In this case, the seat occupancy can be in states $s_i=0, 1, 2$.


Markov state and transition in this situation will be described in Fig.2.
```{r twoseattransition, fig.width = 4}
grViz("
  digraph dot{
    graph[
      rankdir = LR,
      label = 'Fig.2: The Simplest Case(two seats). \nRed means dispatch',
      fontsize = 10,
      layout = neato,
      ordering = out,
    ];
    node[
      shape = circle,
      fixedsize = true
    ];
    a0[label = '0'];
    a1[label = '1'];
    a2[label = '2'];
    a0 -> a1[label = '<I>p</I>@_{1}'];
    a0 -> a2[label = '<I>p</I>@_{2}'];
    a1 -> a2[label = '<I>p</I>@_{1}'];
    a1 -> a2[label = '<I>p</I>@_{2}', color = red];
    a2 -> a1[label = '<I>p</I>@_{1}', color = red];
    a2 -> a2[label = '<I>p</I>@_{2}', color = red];
  }
#  digraph dot{
#    graph[
#      rankdir = LR,
#      label = 'Fig.2: Simplest Case(two seats). \nRed means dispatch',
#      fontsize = 10,
#      layout = dot,
#      ordering = out,
#    ];
#    node[
#      shape = circle,
#      fixedsize = true
#    ];
#    a0[label = '<I>a@_{i}</I>=0'];
#    a1[label = '<I>a@_{i}</I>=1'];
#    a2[label = '<I>a@_{i}</I>=2'];
#    b0[label = '<I>a@_{i+1}</I>=0'];
#    b1[label = '<I>a@_{i+1}</I>=1'];
#    b2[label = '<I>a@_{i+1}</I>=2'];
#    a0 -> b1[label = '<I>p</I>@_{1}'];
#    a0 -> b2[label = '<I>p</I>@_{2}'];
#    a1 -> b2[label = '<I>p</I>@_{1}'];
#    a1 -> b2[label = '<I>p</I>@_{2}', color = red];
#    a2 -> b2[label = '<I>p</I>@_{2}', color = red];
#    a2 -> b1[label = '<I>p</I>@_{1}', color = red];
#    {rank = min; a0;a1;a2;}
#    {rank = max; b0;b1;b2;}
#  } 
")
```
As an element $T_{jk}$ means probability to move from state $s_i=j$ to state $s_{i+1}=k$, we can define transition matrix
$$
T=\begin{pmatrix}
  0 & p_1 & p_2 \\
  0 & 0 & 1\\
  0 & p_1 & p_2
  \end{pmatrix},
$$
since $p_1+p_2=1$.  

#### Stationary State
Let $\mathbf{p}^{(i)}$ denote the state probability distribution after $i$ transitions from an arbitrary initial distribution.
The evolution of the probability distribution follows:
$$
\mathbf{p}^{(i+1)} = \mathbf{p}^{(i)} T
$$
As $i \to \infty$, the distribution converges to the stationary distribution $\pi$, which satisfies:
$$
\pi = \pi T
$$
This means the stationary state $\pi$ is the left eigenvector corresponding to eigenvalue 1 of $T$.
We obtain the stationary state by calculating this eigenvector of $T$.

#### Expectation value of seat occupancy

## Actual setting: one line of four seats
As an example, we calculate seat occupancy in a situation of

- one line of seats
- four seats in the line

This situation matches the seats configuration of Harry Potter and the Forbidden Journey.

### Transition Matrix 
With only a single row, guest must be seated in arrival order.
Therefore, the grouping strategy is determined automatically.

$$
T=\begin{pmatrix}
  p_4 & p_3+p_4& p_2+p_4&p_1+p_4\\
  p_1&0&0&0\\
  p_2&p_1&0&p_2\\
  p_3&p_2&p_1+p_3&p_3
  \end{pmatrix}
$$
 
### Single Queue   

To approach this problem, we introduce a single-parameter model by setting $p_1=p$ and $p_2=p_3=p_4=(1-p)/3$.
This represents the simplest scenario where single guests arrive with probability $p$, and all multi-person groups (sizes 2, 3, 4) are equally likely, each with probability $(1-p)/3$.
Then by an equation $T\pi=\pi$, we can obtain the stationary state
$$
\pi = \frac{1}{6(p+1)(2p^2-p+2)}\begin{pmatrix}
  2p^2+2p+5\\
  2p^3+2p^2+5p\\
  4p^3+4p^2-2p+3\\
  6p^3-2p^2+p+4
  \end{pmatrix}
$$

The dispatch probability is
$$
p_D=\frac{-28p^4+24p^3+6p^2-8p+33}{18(p+1)(2p^2-p+2)}
$$

The expected occupancy rate is
$$
\begin{eqnarray}
\mathrm E[Ocp] &=& \frac{1}{p_D}\frac{3-2p}{4}\\
  &=& \frac{9(p+1)(3-2p)(2p^2-p+2)}{2(-28p^4+24p^3+6p^2-8p+33)}
  \end{eqnarray}
$$


We substitute concrete values to the parameters.
In case of $p=1/20$, 
$$
\pi = \begin{pmatrix}\pi_0\\\pi_1\\\pi_2\\\pi_3\end{pmatrix}= \begin{pmatrix}0.415\\ 0.021\\ 0.236\\ 0.329\end{pmatrix}
$$
and the dispatch transition matrix is
$$
T_D = \begin{pmatrix}
  p_4&p_3+2p_4&p_2+2p_4&p_1+2p_4\\
  0&0&0&0\\
  0&0&0&p_2\\
  0&0&p_3&p_3
  \end{pmatrix}
$$
thus
$$
p_D = \mathbf 1^\mathrm{T}T\pi=0.8827
$$
Therefore, the expected occupancy is
$$
\mathrm E[Ocp] = \begin{pmatrix}
  p_4&p_3+\frac54p_4&p_2+\frac12p_3+\frac32p_4&p_1+\frac34p_2+\frac34p_3+\frac74p_4\\
\end{pmatrix}\begin{pmatrix}\pi_0\\\pi_1\\\pi_2\\\pi_3\end{pmatrix}
  =0.8213
$$

### With Single-Rider Queue
If a single-rider queue is available, the parameters should be modified as
$$
p_1=0, p_2=p_3=p_4=1/3
$$
Then we can obtain the stationary state as
$$
\pi=\begin{pmatrix}0.417\\0\\0.25\\0.333\end{pmatrix}
$$
Using this, we can easily calculate the dispatch probability
$$
D = 0.9167
$$
and the expectation of occupancy
$$
\mathrm E[Ocp]=0.818
$$

The above analysis excludes single riders from the main queue.
However, single riders can fill remaining empty seats, increasing effective occupancy.
If the proportion of available single riders is $p_s$ (assumed to be 0.05), the adjusted expected occupancy becomes:
$$
\mathrm E[Ocp_S]=\mathrm E[Ocp](1+p_s)=0.859
$$

### Pre-grouping
We pre-group guests by merging groups of size 2 or 4 together, keeping groups of size 3 separate.

If group size is only 3, expected occupancy will be 0.75.

In case of group size 2 or 4, we set each probability as $p, 1-p$.

Transition matrix
$$
T = \begin{pmatrix} 1-p&1\\p&0\end{pmatrix}
$$

Stationary state
$$
\pi = \frac{1}{1+p}\begin{pmatrix}1\\p\end{pmatrix}
$$

Expected Dispatch
$$
\mathrm E[D]=\frac{-p^2+p+1}{p+1}
$$

The expected Occupancy
$$
\mathrm E[Ocp]=\frac{p^2-p+2}{2(-p^2+p+1)}
$$
In case $p=1/2$, expected occupancy is minimum 0.9.


## Conclusion
- change 
