---
title: "General Framework for Seat Occupancy Expectation"
author:
  name: "Fukumoto Tomoya"
  affiliation: "Operation Planning"
date: "2018-08-20"
output: html_document
#abstract: "hogehoge is important issue for us. We established a mathematical framework to expect seat occupancy of ride attractions. Our model is based on Markov chain model. Since our model is general, we can aplly to any attraction."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(DiagrammeR)
library(ggplot2)
```

## Introduction
One of the most important issue for theme park operation is raising guestcounts of attractions.
Since guestcounts can be decomposed by dispatch count of vehicles and seat occupancy, we usually focus on these two numbers.
The aim of this study is to establish a mathematical framework to expect seat occupancy of ride attractions.
We revealed that seat occupancy is derivable by statistics of group size of arrival guests, the configuration of seats, queue formation and grouping strategy.
This framework will help us to optimize queue formation and grouping strategy.
The model is so general that the framework will be applied to any attractions.

## Modeling and Framework
### Problem
For ride attractions, grouping operation is deeply related to seat occupancy.
Attraction crew forms a group to dispatch a vehicle from arrival guests not to break a friendship.
That crew is called a grouper.
Guests come again and again.
A grouper guides former arrival guest first.
Grouper can not choose a size of friend group to come next and even expect it.

Almost all seat arrangement of vehicle is designed to be a rectangle.
This means that seats are set in a row, and one or more rows are line up.
Grouping operation is to gather people to have a seat in the same row.
And of course friend group should be in the same group and sit next to each other.

Grouping problem is how to maximize guest experience to fill seats with guest group.
Grouper ask a size of group and select a row to guide.
Grouping strategy is to select an optimal row when a size of group come.

Therefore there exists an upper bound of seat occupancy in each attractions.

The upper bound would depend on statistics of group size of arrival guest and a pattern of seats configuration.
If we have a knowledge of upper bound of an attraction, we can evaluate its efficiency of grouping operation and whether the seat occupancy can improve or not.

### Mathematical Modeling
Our problem can be handled as a state-action-reward model.

- **state**:  the seats pattern filled by guests
- **action**: decision which row to guide arrival guests and to dispatch a vehicle or not
- **reward**: the number of guests who take a seat when a vehicle is dispatched

To treat this model, we define a numbers of variables for state and transition

- the size of $t$th arrival group is a probability variable $x_t \in \mathbb N$.
- the seats filled pattern after the $t$th group has seats $s_t$.
- the set of pattern of seats filled $\sigma \in \Sigma$.

After a vehicle is dispatched, a new vacant vehicle will come.
This means the state will be initialized as $\sigma_0$.

```{r markov, fig.width = 5, fig.height = 2}
grViz(
  "digraph dot{
    graph[
      rankdir = LR,
      label = 'Fig.1: State, Action, and Transition',
      fontsize = 15
    ]
    node[
      fixedsize = false
    ]
      a1[label = '<I>s</I>@_{<I>t</I>}=<I>&sigma;</I>', shape = circle];
      a2[label = 'Which row\n to guid?', shape = square];
    edge[label = '<I>x@_{t}</I>'] a1 -> a2 ;
      a3[label = 'Dispatch\n or not?', shape = square];
    edge[label = ''] a2 -> a3;
      a4[label = '<I>s</I>@_{<I>t</I>+1}=<I>&sigma;</I>&prime;', shape = circle];
      edge[label = ''] a3 -> a4;

}
  "
)
```

We assume that the situation is stationary.
The assumption induces the probability that a size $k$ group arrives is independent on time $P[x_t=k]=p_k$ .

## Simple case: Markov chain
If the seat layout is one line, then our option is giving seats or dispatch a vehicle because we can not choose which seats we provide.
This means that statistics of a guest arrival directly gives the probability of transition from a state to another.
We can handle such problems as a simple Markov chain model.

In this case, the number of occupied seats describes the state completely,

$$
\sigma \in \Sigma = \{1,2,\ldots,N\}
$$

where $N$ is the number of seats in the line.
We can calculate the probability being each state $P[s_t=k]$ from $p_k$, and the expectation of seat occupancy from the state probabilities.
This expectation value should be the upper bound of seat occupancy.

### Steps to solve the problem
We 

1. Give a transition matrix
2. Identify the stationary state.
3. Calculate the expectation value of occupancy from stationary probability.

#### transition matrix
Suppose that a group sized $k$ has arrived.
In case the number of empty seats is greater than $k$, they will have their seats.
If it's less than $k$, the crew dispatch the vehicle and give them a new vacant vehicle.

- $s_t<x_t$ then $s_{t+1}=s_t+x_t$ 
- $s_t>x_t$ then $s_{t+1}=x_t$

Because arrival probability is independent on state, we can calculate probabilities from state $j$ to $k$ as
$$
T_{jk}=P[s_{t+1}=k|s_{t}=j]=(p_{j-k}+p_k)P[s_{t}=j]
$$
where $p_k=0$ for $k\leq0$.
And we call a matrix which elements are these probabilities as **transition matrix**.

We show an example in the simplest case, which has only two seats.
In this case, seat occupancy can take $s_t=0, 1, 2$.

Markov state and transition in this situation can be shown in Fig.2.
```{r twoseattransition, fig.width = 4}
grViz("
  digraph dot{
    graph[
      rankdir = LR,
      label = 'Fig.2: The Simplest Case(two seats). \nRed means dispatch',
      fontsize = 10,
      layout = neato,
      ordering = out,
    ];
    node[
      shape = circle,
      fixedsize = true
    ];
    a0[label = '0'];
    a1[label = '1'];
    a2[label = '2'];
    a0 -> a1[label = '<I>p</I>@_{1}'];
    a0 -> a2[label = '<I>p</I>@_{2}'];
    a1 -> a2[label = '<I>p</I>@_{1}'];
    a1 -> a2[label = '<I>p</I>@_{2}', color = red];
    a2 -> a1[label = '<I>p</I>@_{1}', color = red];
    a2 -> a2[label = '<I>p</I>@_{2}', color = red];
  }
  ")
```
In this case our transition matrix is
$$
T=\begin{pmatrix}
  0 & p_1 & p_2 \\
  0 & 0 & 1\\
  0 & p_1 & p_2
  \end{pmatrix},
$$
since $p_1+p_2=1$.  

#### Stationary State
If state probability is given by a vector $A_t=(P[s_t=0],P[s_t=1],P[s_t=2])$, then next step $A_{t+1}=A_tT$.
We can calculate a state probability in each step.
If the calculation converges, state probability satisfies an equation
$$
A=AT
$$
This probability vector is called a **stationary distribution**.
The stationary distribution is an eigenvector corresponding eigenvalues 1 of $T$.

We empirically know that it is very fast to approach the stationary distribution.
So we can expect we observe the states with the stationary distribution in most of the time.
This means that the expectation of seat occupancy from the stationary distribution is a reasonable estimation.

By easy calculation, we can get the stationary distribution as
$$
A = \left(0, \frac{p}{p+2}, \frac{2}{p+2}\right)
$$
where $p=p_1=1-p_2$.

#### Expectation value of seat occupancy
The expectation of seat occupancy can be described with the conditional probability
$$
\mathrm E\left[ \mathrm{ Seat Occupancy } | \mathrm{ Dispatch } \right] = \frac{\mathrm E[\mathrm{OccupancyWhenDispatch}]}{P[\mathrm{Dispatch}]}.
$$

The probability of dispatch is
$$
P[\mathrm{Dispatch}]=a_1p_2+a_2(p_1+p_2) = \frac{-p^2+p+2}{p+2}
$$

And the numerator is
$$
\mathrm{E}[\mathrm{OccupancyWhenDispatch}]=1\cdot a_1p_2 + 2\cdot a_2(p_1+p_2) = \frac{-p^2+p+4}{p+2}
$$

Thus our goal is
$$
\mathrm E\left[ \mathrm{ Seat Occupancy } | \mathrm { Dispatch } \right] = \frac{-p^2+p+4}{-p^2+p+2} = 2-\frac{p(1-p)}{p(1-p)+2}.
$$

```{r simplest_occupancy}
p <- seq(0, 1, by = 0.01)
fun <- function(p) return((-p^2+p+4)/(-p^2+p+2))
df <- data.frame(p = p, seat_occupancy = fun(p))
ggplot(df) +
  geom_line(aes(p, seat_occupancy)) +
  scale_y_continuous(breaks = seq(1.85,2,by=0.05)) +
  labs(
       title = "Fig3. Seat Occupancy vs Arrival Probability of single guest",
       subtitle = "uncertainity reduces occupancy expectation",
       y = "Occupancy Expectation"
  )
```

We can see the expectation of seat occupancy has the greatest value 2 in $p=0,p=1$.
This is obvious because in the case the arrival group size is fixed with 1 or 2.
In the most uncertain case $p=1/2$ our expectation is the least $17/9\sim1.89$

We completely solved the problem.
Logically we could solve any other situation having one line seats in the same manner.
But it does not mean we can solve such problem easily.

## Actual setting: one line of four seats
As an example, we consider a situation

- one line of seats 
- four seats in a line

This corresponds the seats configuration of the Harry Potter and the Forbidden Journey.

Because we have only one line, we have to provide guests in arrival order.
Thus the expectation of seat occupancy can be evaluated with Markov chain model.

### Occupancy expectation
The transition of state is described by below.

```{r FBJtransition, fig.width = 8}
grViz("
  digraph dot{
    graph[
      label = 'Fig.4: One line, four seats. Red means dispatch. The state 0 is removed',
      fontsize = 10
    ];
    node[
      shape = circle,
      fixedsize = true,
      fontsize = 10
    ];
    edge[
	 fontsize = 10,
	 arrowsize = 0.5
    ];
    a1[label = '1'];
    a2[label = '2'];
    a3[label = '3'];
    a4[label = '4'];
    a1 -> a2[label = 'p@_{1}'];
    a1 -> a3[label = 'p@_{2}'];
    a1 -> a4[label = 'p@_{3}'];
    a1 -> a4[label = 'p@_{4}', color = 'red'];
    a2 -> a3[label = 'p@_{1}'];
    a2 -> a4[label = 'p@_{2}'];
    a2 -> a3[label = 'p@_{3}', color = 'red'];
    a2 -> a4[label = 'p@_{4}', color = 'red'];
    a3 -> a4[label = 'p@_{1}'];
    a3 -> a2[label = 'p@_{2}', color = 'red'];
    a3 -> a3[label = 'p@_{3}', color = 'red'];
    a3 -> a4[label = 'p@_{4}', color = 'red'];
    a4 -> a1[label = 'p@_{1}', color = 'red'];
    a4 -> a2[label = 'p@_{2}', color = 'red'];
    a4 -> a3[label = 'p@_{3}', color = 'red'];
    a4 -> a4[label = 'p@_{4}', color = 'red'];
  }
  ")
```

and transition matrix is

$$
T=\begin{pmatrix}
  0 & p_1 & p_2 & p_3+p_4\\
  0 & 0   & p_1+p_3 & p_2+p_4\\
  0 & p_2 & p_3 & p_1 + p_4\\
  p_1 & p_2 & p_3 & p_4\\
  \end{pmatrix}.
$$

Then we obtain the stationary distribution $\pi=(\pi_i)$ with calculation

$$
\pi = \begin{pmatrix}\pi_1\\\pi_2\\\pi_3\\\pi_4\\\end{pmatrix} = 
\begin{pmatrix}
p_1(1-p_1p_2-p_2p_3-p_3)\\
p_1p_2^2+p_2+p_1^2-p_1^2p_3\\
2p_1p_2+p_1^3+p_1^2p_3+p_2p_3+p_3\\
1-p_1p_2-p_2p_3-p_3\\
\end{pmatrix} / Z
$$

where 
$$
Z = p_1(1-p_1p_2-p_2p_3-p_3+p_2^2+p_1+p_2+p_1^2) + p_2+1.
$$


Then we obtain the occupancy expectation
$$
\mathrm E[\mathrm{Occupancy}] = \frac{\mathrm{E}[\mathrm{OccupancyWhenDispatch}]}{P[\mathrm{Dispatch}]}
$$
where
$$
\mathrm{E}[\mathrm{OccupancyWhenDispatch}] = \pi_1p_4+2\pi_2(p_3+p_4)+3\pi_3(p_2+p_3+p_4)+4\pi_4(p_1+p_2+p_3+p_4)
$$
$$
P[\mathrm{Dispatch}] = \pi_1p_4+\pi_2(p_3+p_4)+\pi_3(p_2+p_3+p_4)+\pi_4(p_1+p_2+p_3+p_4)
$$

Because this value is continious, there is the maximum and minimum value in $p_i\in[0,1],\sum p_i=1$.
We would obtain the value using the Lagrange multiplier.
 
### With Single-Rider Queue
We suppose configuration of a single-rider queue.
A single-rider queue is an independent queue only for guests of group size 1.
In this configuration, we evaluate occupancy expectation without the arrivals of group size 1, and adjust its effect later.
We use arrival probabilities given no single guests
$$
p_2' = P[s_t=2|s_t\neq1]=p_2/(1-p_1)\\
p_3' = P[s_t=3|s_t\neq1]=p_3/(1-p_1)\\
p_4' = P[s_t=4|s_t\neq1]=p_4/(1-p_1).
$$

The stationary distribution can be simplified much in case $p_1=0$,
$$
\pi = \begin{pmatrix}0\\\frac{p_2'}{1+p_2'}\\p_3'\\\frac{p_2'^2}{1+p_2'}+p_4'\\\end{pmatrix}.
$$
$$
\mathrm{E}[Ocp|s_t\neq1] = \frac{2p_2'+3p_3'+4p_4'}{\frac{p_2'}{1+p_2'}+p_3'+p_4'}
$$

Actually we have guests of group size 1 in $p_1$.
The ratio of the single guests in the total guest count can be evaluated
$$
r = \frac{p_1}{p_1+2p_2+3p_3+4p_4} \cdots (1)
$$

Single guests will take their seat if a vacant seat is left before dispatch.
Thus we just add them to occupancy expectation.

$$
\mathrm{E}[Ocp Single] = \mathrm{E}[Ocp|s_t\neq1](1+r) = \frac{2p_2+3p_3+4p_4}{\frac{(1-p_1)p_2}{1-p_1+p_2}+p_3+p_4}\frac{2p_1+2p_2+3p_3+4p_4}{p_1+2p_2+3p_3+4p_4}
$$

To compare with the original the one normal queue and queues having isolated single rider queue, we assume arrival probability
$$
\begin{pmatrix}p_1\\p_2\\p_3\\p_4\end{pmatrix}=\begin{pmatrix}1/10\\1/3\\1/3\\1/3\end{pmatrix} \cdots (2)
$$
In the case,
```{r compare_single}
stat_dist <- function(p1,p2,p3,p4){
  z <- p1*(1-p1*p2-p2*p3-p3+p2^2+p1+p2+p1^2)+p2+1
  pi1 <- p1*(1-p1*p2-p2*p3-p3)/z
  pi2 <- (p1*p2^2+p2+p1^2-p1^2*p3)/z
  pi3 <- (2*p1*p2+p1^3+p1^2*p3+p2*p3+p3)/z
  pi4 <- (1-p1*p2-p2*p3-p3)
  return(c(pi1,pi2,pi3,pi4))
}
ocp <- function(p1,p2,p3,p4){
  pi1 <- stat_dist(p1,p2,p3,p4)[1]
  pi2 <- stat_dist(p1,p2,p3,p4)[2]
  pi3 <- stat_dist(p1,p2,p3,p4)[3]
  pi4 <- stat_dist(p1,p2,p3,p4)[4]
  numerator <- pi1*p4 + 2*pi2*(p3+p4) + 3*pi3*(p2+p3+p4)+4*pi4
  denominator <- pi1*p4 + pi2*(p3+p4) + pi3*(p2+p3+p4) + pi4
  return(numerator/denominator)
}
eocp <- ocp(1/10,1/3,1/3,1/3)
eocp_sg <- ocp(0,3/10,3/10,3/10)*(2*1/10+2/3+3/3+4/3)/(1/10+2/3+3/3+4/3)
```
$$
E[Ocp] = `r eocp`\\
E[OcpSingle] = `r eocp_sg`
$$

We can expect `r eocp_sg-eocp` improvement of occupancy by using the single rider queue.

#### Knowledge: minimum occupancy without single arrival
In case of $p_1=0$, what is the lowest value of the occupancy expectation for $p_2,p_3,p_4$? 
The values of variables should be extremum,
$$
\frac{\partial}{\partial p_2}\mathrm E[Ocp|s_t\neq1]=0\\
\frac{\partial}{\partial p_3}\mathrm E[Ocp|s_t\neq1]=0\\
\frac{\partial}{\partial p_4}\mathrm E[Ocp|s_t\neq1]=0.
$$
But the solution of equations does not exist in $p_2,p_3,p_4\in[0,1], p_2+p_3+p_4=1$.
Thus a boundary must be our goal.

Comparing with boundaries, we got the minimum value $\frac{4\sqrt{5}+16}{9}\sim2.77$ at $(p_2,p_3,p_4)=(\frac{\sqrt{5}-1}{2},\frac{3-\sqrt{5}}{2},0)$ which is the solution of
$$
\frac{\partial}{\partial p_3}\mathrm{E}[Ocp|p_1=0, p_4=0] = 0.
$$

#### Question
When we configure two queues, is it optimal to separate singles and the others? (Maybe yes)

### Pre-grouping
If the facility has redundancy, we can make one more queue line.
We distribute arrival guests into different queue lines by its group size.
The operation is called pre-grouping.

We hope to know

- Does pre-grouping improve seat occupancy?
- How much is its improvement?
- What is the best pre-grouping strategy?

When we suppose to use a single-rider queue, our pre-grouping option is below

1. Queue1:$x_t\in\{2,3\}$, Queue2: $x_t=4$
2. Queue2:$x_t\in\{2,4\}$, Queue2: $x_t=3$
3. Queue3:$x_t\in\{3,4\}$, Queue3: $x_t=2$

Which is the best pre-grouping?
We use actual arrival probabilities assumed before (2) $(p_i)=(1/10,1/3,1/3,1/3)$ to compare with them.

#### Option 1 (2,3)-4
For queue 1, arrival probabilities is transformed
$$
p_2' = \frac{p_2}{p_2+p_3}\\
p_3' = \frac{p_3}{p_2+p_3}
$$
and the occupancy expectation would be
$$
\mathrm {E[Occupancy|Option1,Queue1]} = \frac{2p_2'+3p_3'}{\frac{p_2'}{1+p_2'}+p_3}
$$
For queue 2, we only receive group size 4 therefore $\mathrm{E[Ocp|O1,Q2]}=4$.


We combine them with arrival guest count ratio.
$$
\mathrm{E[Occupancy|Option1]} = \left(\frac{2p_2+3p_3}{p_1+2p_2+3p_3+4p_4}\frac{2p_2'+3p_3'}{\frac{p_2'}{1+p_2'}+p_3'} + \frac{4p_4}{p_1+2p_2+3p_3+4p_4}\cdot 4 \right)(1+r)
$$
where $r$ is a single rider effect (1).

```{r exocp_pg_o1}
exocp_pg_o1 <- function(p1,p2,p3,p4){
  exg <- p1 + 2*p2 + 3*p3 + 4*p4
  e23 <- (2*p2 + 3*p3)/exg * (2*p2 + 3*p3)/(p2*(p2 + p3)/(2*p2 + p3) + p3)
  e4 <- 4*p4/exg*4
  r <- p1/exg
  return( (e23+e4)*(1+r))
}
```

We substitute actual probability values (2) and obtain $\mathrm {E[Ocp|O1]}= `r exocp_pg_o1(1/10,1/3,1/3,1/3)`$ .

#### Option 2 (2,4)-3
For queue 1
$$
p_2' = \frac{p_2}{p_2+p_4}\\
p_4' = \frac{p_4}{p_2+p_4}\\
\mathrm {E[Occupancy|Option2,Queue1]} = \frac{2p_2'+4p_4'}{\frac{p_2'}{1+p_2'}+p_4}.
$$
Then
$$
\mathrm{E[Occupancy|Option2]} = \left(\frac{2p_2+4p_4}{p_1+2p_2+3p_3+4p_4}\frac{2p_2'+4p_4'}{\frac{p_2'}{1+p_2'}+p_4'} + \frac{3p_3}{p_1+2p_2+3p_3+4p_4}\cdot 3 \right)(1+r)
$$

```{r exocp_pg_o2}
exocp_pg_o2 <- function(p1,p2,p3,p4){
  exg <- p1 + 2*p2 + 3*p3 + 4*p4
  e24 <- (2*p2 + 4*p4)/exg * (2*p2 + 4*p4)/(p2*(p2 + p4)/(2*p2 + p4) + p4)
  e3 <- 3*p3/exg*3
  r <- p1/exg
  return( (e24+e3)*(1+r))
}
```

If we substitute actual probability values (2) and obtain $\mathrm {E[Ocp|O2]}= `r exocp_pg_o2(1/10,1/3,1/3,1/3)`$ .

#### Option 3 (3,4)-2
For queue 1
$$
p_3' = \frac{p_3}{p_3+p_4}\\
p_4' = \frac{p_4}{p_3+p_4}\\
\mathrm {E[Occupancy|Option3,Queue1]} = \frac{3p_3'+4p_4'}{p_3'+p_4}.
$$
Then
$$
\mathrm{E[Occupancy|Option3]} = \left(\frac{3p_3+4p_4}{p_1+2p_2+3p_3+4p_4}\frac{3p_3'+4p_4'}{p_3'+p_4'} + \frac{2p_2}{p_1+2p_2+3p_3+4p_4}\cdot 4 \right)(1+r)
$$

```{r exocp_pg_o3}
exocp_pg_o3 <- function(p1,p2,p3,p4){
  exg <- p1 + 2*p2 + 3*p3 + 4*p4
  e34 <- (3*p3 + 4*p4)/exg * (3*p3 + 4*p4)/(p3 + p4)
  e2 <- 2*p2/exg*4
  r <- p1/exg
  return( (e34+e2)*(1+r))
}
```

If we substitute actual probability values (2) and obtain $\mathrm {E[Ocp|O3]}= `r exocp_pg_o3(1/10,1/3,1/3,1/3)`$ .

#### Question
If our arrival probability was changed, the best pre-grouping option would be changed.
Which is its boundary?

## Conclusion
We introduce the framework to evaluate seat occupancy expectation from arrival group size statistics, seat configuration and grouping strategy.
And we calculate the expectation in some simple situations by using Markov chain model.
The case of five or more seats in a line would be calculated in the future work.

We only evaluate the expectation.
But more analysis would make more useful knowledge, 
The relationship between the best grouping strategy and arrival probabilities is interesting because arrival probabilities can be changed over timein actual setting and we want to know the stability of our grouping strategy.

The most strong constraint was the one line seat configuration.
In the case, we can not apply Markov chain model.
Thus we need different model to analyze.

