---
title: "General Framework for Seat Occupancy Expectation"
author:
  name: "Fukumoto Tomoya"
  affiliation: "Operation Planning"
date: "2018-08-20"
output: html_document
#abstract: "hogehoge is important issue for us. We established a mathematical framework to expect seat occupancy of ride attractions. Our model is based on Markov chain model. Since our model is general, we can aplly to any attraction."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction
One of the most important issue for park operation is raising guestcounts of attractions.
Since guestcounts can be decomposed by dispatch count and seat occupancy, we focus on these two numbers.
The aim of this study is to establish a mathematical framework to expect seat occupancy of ride attractions.
We revealed that seat occupancy is derivable by statistics of group size of arrival guests, the number of seats, queue formation and grouping strategy.
This framework will help us to optimize queue formation and grouping strategy.
Because the model is general, that is, the model is not dependent on any actual attraction, the framework can be applied to any attractions.

## Modeling and Framework
### Overview
For ride attractions, grouping operation is deeply related to seat occupancy.
Crew create a group to dispatch from arrival guests not to break each group of guests.
Thus if a group size of arrival guest is not suited, crew will not be able to make seat full however efficient they are.
This fact indicates the existence of upper bound of seat occupancy in each attractions.

The upper bound would depend on statistics of group size of arrival guest and the number of seats for a vehicle.
If we have a knowledge of upper bound of an attraction, we can clarify its efficiency of grouping operation and whether the seat occupancy can improve or not by comparing the upper bound and the actual value.
And we will be able to judge if a change of operation can achieve more seat occupnacy.

The upper bound can be probability variable because arrivals of guest should be probabilistic.
Thus we will obtain upper bounds as expectation values.  

### Mathematical Modeling
To evaluate the upper bound, we define few numbers for situation of operation.

- the size of $i$th arrival group is probability variable $x_i$.
- the number of seats filled by guests is a probability variable $a_i$ after the $i$the group have seats.

And we assume that situation is stationary.

- the probability that a size $k$ group arrives is time stationary and $P[x_i=k]=p_k$ .

We apply **Markov chain model** to this problem.
Markov chain model is a kind of stochastic model having state and transition.  
This model focuses relationships between states and transition and a lot of techniques is established to solve this problem.

We consider state after $i$th transition as $a_i$ and the probability of transition as $p_k$.

```{r markov, fig.width = 3, fig.height = 2}
library(DiagrammeR)
grViz(
  "digraph dot{
    graph[
      rankdir = LR,
      label = 'Fig.1: State and Transition',
      fontsize = 10
    ]
    node[
      shape = circle,
      fixedsize = true
    ]
      a1[label = '<I>a@_{i}</I>'];
      a2[label = '<I>a</I>@_{<I>i</I>+1}']
    edge[label = '<I>p@_{k}</I>']
      a1 -> a2}
  "
)
```

Then our problem for expectation of seat occupancies is translated into how to find out transition probability $P[x_i=k]=p_k$ and how to derive the probability having each states $P[a_i]$ from $p_k$ because the expectation value can be easily calculated by $p_k$ and $P[a_i]$.
This expectation value should be the upper bound of seat occupancy.

### Steps to solve the problem
We 

1. Set a transition matrix from seat number and grouping strategy.
2. Identify the stationary state.
3. Calculate expectation value of occupancy from stationary probability.

#### Setting a transition matrix
Suppose that a group sized $k$ has arrived.
In case the number of empty seats is greater than $k$, they will take their seats.
If it's less than $k$, the vehicle will dispatch and the group will take their seats of new vehicle.  

- $a_i<x_i$ then $a_{i+1}=a_i+x_i$ 
- $a_i>x_i$ then $a_{i+1}=x_i$

This means that a probability having a state can be described by previous state and arrival probability.
Because arrival probability is indenpendent on state, these probabilities are 
$$
\begin{eqnarray}
T_{jk}&=&P[a_{i+1}=k|a_{i}=j]\\
  &=&\begin{cases}p_{j-k}P[a_{i}=j]&(j-k>0)\\ p_{k}P[a_{i}=j]&(j-k\leq0)\end{cases}\\
\end{eqnarray}
$$
And we call a matrix which elements are these probabilities as **transition matrix**.

We show an example in the simplest case, which has only two seats.
In this case, seat occupancy can take states from zero to two $a_i=0, 1, 2$.


Markov state and transition in this situation will be described in Fig.2.
```{r twoseattransition, fig.width = 4}
grViz("
  digraph dot{
    graph[
      rankdir = LR,
      label = 'Fig.2: Simplest Case(two seats). \nRed means dispatch',
      fontsize = 10,
      layout = neato,
      ordering = out,
    ];
    node[
      shape = circle,
      fixedsize = true
    ];
    a0[label = '0'];
    a1[label = '1'];
    a2[label = '2'];
    a0 -> a1[label = '<I>p</I>@_{1}'];
    a0 -> a2[label = '<I>p</I>@_{2}'];
    a1 -> a2[label = '<I>p</I>@_{1}'];
    a1 -> a2[label = '<I>p</I>@_{2}', color = red];
    a2 -> a1[label = '<I>p</I>@_{1}', color = red];
    a2 -> a2[label = '<I>p</I>@_{2}', color = red];
  }
#  digraph dot{
#    graph[
#      rankdir = LR,
#      label = 'Fig.2: Simplest Case(two seats). \nRed means dispatch',
#      fontsize = 10,
#      layout = dot,
#      ordering = out,
#    ];
#    node[
#      shape = circle,
#      fixedsize = true
#    ];
#    a0[label = '<I>a@_{i}</I>=0'];
#    a1[label = '<I>a@_{i}</I>=1'];
#    a2[label = '<I>a@_{i}</I>=2'];
#    b0[label = '<I>a@_{i+1}</I>=0'];
#    b1[label = '<I>a@_{i+1}</I>=1'];
#    b2[label = '<I>a@_{i+1}</I>=2'];
#    a0 -> b1[label = '<I>p</I>@_{1}'];
#    a0 -> b2[label = '<I>p</I>@_{2}'];
#    a1 -> b2[label = '<I>p</I>@_{1}'];
#    a1 -> b2[label = '<I>p</I>@_{2}', color = red];
#    a2 -> b2[label = '<I>p</I>@_{2}', color = red];
#    a2 -> b1[label = '<I>p</I>@_{1}', color = red];
#    {rank = min; a0;a1;a2;}
#    {rank = max; b0;b1;b2;}
#  } 
")
```
As an element $T_{jk}$ means probability to move from state $a_i=j$ to state $a_{i+1}=k$, we can define transition matrix
$$
T=\begin{pmatrix}
  0 & p_1 & p_2 \\
  0 & 0 & 1\\
  0 & p_1 & p_2
  \end{pmatrix},
$$
since $p_1+p_2=1$.  

#### Stationary State
Set state probability
$$
A_i = \begin{pmatrix}a\\b\\c\end{pmatrix}
$$
next step $A_{i+1}=A_iT$.
we can calculate state probability in each step.
if we continue to calculation and state converge, state probability should be sufficient equation
$$
A=AT
$$
This means stationary state is eigenvector corresponding eigenvalues 1 of $T$.
So we obtain stationary state by calculating eigenvector of $T$.

Example:

#### Expectation value of seat occupancy

## Example: Single Grouping
As an example, we calculate seat occupancy in a situation of

- one line of seats
- four seats in the line

This situation is fit attractions like Harry Potter and the Forbidden Journey.

We 

### Transition Matrix 
If we make only one line, we have to group guests in arrival order.
Thus our grouping strategy can be decided automatically.
$$
T=\begin{pmatrix}
  p_4 & p_3+p_4& p_2+p_4&p_1+p_4\\
  p_1&0&0&0\\
  p_2&p_1&0&p_2\\
  p_3&p_2&p_1+p_3&p_3
  \end{pmatrix}
$$
 
### Single Queue   

To approach this problem, we assume $p_1=p, p_2=p_3=p_4=q/3$.
This is the simplest model which has only two parameters, single guest ratio and grouped guest.
And each grouped guests come to an attraction in the same possibility.
Then by an equation $Tx=x$, we can obtain the stationary state
$$
x = \frac{1}{6(p+1)(2p^2-p+2)}\begin{pmatrix}
  2p^2+2p+5\\
  2p^3+2p^2+5p\\
  4p^3+4p^2-2p+3\\
  6p^3-2p^2+p+4
  \end{pmatrix}
$$

The probability of dispatch
$$
p_D=\frac{-28p^4+24p^3+6p^2-8p+33}{18(p+1)(2p^2-p+2)}
$$

Expected occupancy
$$
\begin{eqnarray}
\mathrm E[Ocp] &=& \frac{1}{p_D}\frac{3-2p}{4}\\
  &=& \frac{9(p+1)(3-2p)(2p^2-p+2)}{2(-28p^4+24p^3+6p^2-8p+33)}
  \end{eqnarray}
$$


We substitute concrete values to the parameters.
In case of $p=1/20$, 
$$
\pi = \begin{pmatrix}\pi_0\\\pi_1\\\pi_2\\\pi_3\end{pmatrix}= \begin{pmatrix}0.415\\ 0.021\\ 0.236\\ 0.329\end{pmatrix}
$$
and dispatch transition is
$$
T_D = \begin{pmatrix}
  p_4&p_3+2p_4&p_2+2p_4&p_1+2p_4\\
  0&0&0&0\\
  0&0&0&p_2\\\
  0&0&p_3&p_3
  \end{pmatrix}
$$
thus
$$
p_D = \mathbf 1^\mathrm{T}T\pi=0.8827
$$
Then we obtain the expected occupancy is 
$$
\mathrm E[Ocp] = \begin{pmatrix}
  p_4&p_3+\frac54p_4&p_2+\frac12p_3+\frac32p_4&p_1+\frac34p_2+\frac34p_3+\frac74p_4\\
\end{pmatrix}\begin{pmatrix}\pi_0\\\pi_1\\\pi_2\\\pi_3\end{pmatrix}
  =0.8213
$$

### With Single Rider Queue
If we have a single rider queue, the parameters should be modified as
$$
p_1=0, p_2=p_3=p_4=1/3
$$
Then we can obtain the stationary state as
$$
\pi=\begin{pmatrix}0.417\\0\\0.25\\0.333\end{pmatrix}
$$
Using this, we can easily calculate dispatch probability
$$
D = 0.9167
$$
and the expectation of occupancy
$$
\mathrm E[Ocp]=0.818
$$
Finaly, we adjust a single rider effect as 
$$
\mathrm E[Ocp_S]=\mathrm E[Ocp](1+p_s)=0.859
$$
where $p_s$ is a single rider availability assumed by 0.05.

### Pre-grouping
- we separate guests by its group size (2,4) and 3 before grouping

If group size is only 3, expected occupancy will be 0.75.

In case of group size 2 or 4, we set each probability as $p, 1-p$.

Transition matrix
$$
T = \begin{pmatrix} 1-p&1\\p&0\end{pmatrix}
$$

Stationary state
$$
\pi = \frac{1}{1+p}\begin{pmatrix}1\\p\end{pmatrix}
$$

Expected Dispatch
$$
\mathrm E[D]=\frac{-p^2+p+1}{p+1}
$$

Expected Occupancy
$$
\mathrm E[Ocp]=\frac{p^2-p+2}{2(-p^2+p+1)}
$$
In case $p=1/2$, expected occupancy is minimum 0.9.


## Conclusion
- change 
